---
title: "Survival pathways"
output: html_notebook
---

```{r warning=FALSE, message=FALSE}
library(dplyr) # loads dplyr
library(survival) # core survival analysis functions
library(survminer) # recommended for visualizing survival curves
# Bioconductor packages
library(RTCGA)
library(RTCGA.clinical)
library(RTCGA.mRNA)
```


```{r warning=FALSE, message=FALSE}
#library(DESeq2)
#library(ggplot2)
#library(vsn)
#library(pheatmap)
#library(ggrepel)
set.seed(1)
```

# Собираем данные

clinical data

```{r}
clinkLung <- survivalTCGA(LUSC.clinical,
                          extract.cols=c("admin.disease_code",
                                         "patient.gender",
                                         "patient.days_to_death",
                                         "patient.days_to_last_followup"))

head(clinkLung)

hist(clinkLung$times)
```

RNA

```{r}
raw.counts <- readRDS("LUSC_counts.rds")
dim(raw.counts)
as.data.frame(raw.counts[1:5, 1:5])
```
```{r}
dim(raw.counts)
counts.rna <- raw.counts[,!duplicated(substr(colnames(raw.counts), 1, 12))]
dim(counts.rna)
colnames(counts.rna) <- substr(colnames(counts.rna), 1, 12)
patients.rows <- Reduce(intersect,list(clinkLung$bcr_patient_barcode,colnames(counts.rna)))
counts.rna <- counts.rna %>% 
  as.data.frame() %>%
  dplyr::select(order(colnames(counts.rna))) %>%
  dplyr::select(patients.rows) %>%
  dplyr::slice_max(rowSums(.), n = 10000)
dim(counts.rna)
counts.rna[1:5, 1:5]
```

для cox regression надо, чтобы данные были в лог-шкале и нормализованные между образцами
нормализацию можно через limma::voom, можно через rlog (но это вероятно будет долго), можно edgeR TMM и log после этого

```{r}
counts.rna.voom <- limma::voom(counts.rna, plot = T)

head(as.data.frame(counts.rna.voom$targets))
head(as.data.frame(counts.rna.voom$weights))
head(as.data.frame(counts.rna.voom$design))
#head(as.data.frame(counts.rna.voom$E))
boxplot(as.data.frame(counts.rna.voom[, 1:10]$E))

#counts.rna.log.norm <- limma::normalizeBetweenArrays(log2(counts.rna + 1))
#boxplot(counts.rna.log.norm[, 1:10])
```

concatenate rna + clinical data

```{r}
clink.and.rna <- as.data.frame(t(counts.rna.voom$E))
clink.and.rna$bcr_patient_barcode <- substr(rownames(clink.and.rna), 1, 12)
clink.and.rna <- clink.and.rna %>%
  inner_join(clinkLung, by="bcr_patient_barcode")
dim(clink.and.rna)
head(clink.and.rna)
```

cox regression для каждого гена

```{r}
Survfun <- function(gene) as.formula(paste("Surv(times,patient.vital_status)",gene,sep="~"))

survival.coxph <- sapply(setdiff(colnames(clink.and.rna),colnames(clinkLung)),
                         function(gene) summary(coxph(Survfun(gene), data=clink.and.rna)))
```

```{r}
survival.coxph <- survival.coxph %>% t() %>% as.data.frame() %>% select(coefficients)
survival.coxph[1,]
coxph(Surv(times,patient.vital_status)~MT.CO1, data=clink.and.rna)
```

The exp(coef) column contains eβ1
 (see background section above for more info). This is the hazard ratio – the multiplicative effect of that variable on the hazard rate (for each unit increase in that variable).

HR=1: No effect
HR>1: Increase in hazard
HR<1: Reduction in hazard (protective)

The p-value comes from testing the null hypothesis that this hazard ratio is 1, or that there is no difference in the relative risk of the event comparing individuals with varying levels of LVEF. When you control for multiple covariates at the same time, the interpretation of the hazard ratio changes somewhat.

задача - отранжировать все экспрессируемые гены по их влиянию на выживаемость

```{r}
survival.coxph.rownames <- rownames(survival.coxph)

survival.coxph <- as.data.frame(matrix(unlist(survival.coxph),ncol=5,byrow=TRUE))
rownames(survival.coxph) <- survival.coxph.rownames
colnames(survival.coxph) <- c("coef", "HR", "se", "z", "Pval")

head(survival.coxph %>% arrange(Pval))
```

```{r}
clink.and.rna.amounts <- clink.and.rna
```

```{r}
mean(clink.and.rna$FGA)
hist(clink.and.rna$FGA)
ggplot(clink.and.rna, aes(FGA)) + geom_histogram(bins=20)
```

```{r}
clink.and.rna.amounts$FGA_amount <-
  cut(clink.and.rna.amounts$FGA,
      breaks=c(-Inf, -0.225, Inf),
      labels=c("low", "high"))

ggsurvplot(survfit(Surv(times, patient.vital_status)~FGA_amount,
                   data=clink.and.rna.amounts), pval=TRUE, conf.int=TRUE)
```

```{r}
mean(clink.and.rna$MRPS21)
hist(clink.and.rna$MRPS21)
ggplot(clink.and.rna, aes(MRPS21)) + geom_histogram(bins=20)
```

```{r}
clink.and.rna.amounts$MRPS21_amount <-
  cut(clink.and.rna.amounts$MRPS21,
      breaks=c(-Inf, 5.8, Inf),
      labels=c("low", "high"))

ggsurvplot(survfit(Surv(times, patient.vital_status)~MRPS21_amount,
                   data=clink.and.rna.amounts), pval=TRUE, conf.int=TRUE)
```

```{r}
mean(clink.and.rna$SLFN13)
hist(clink.and.rna$SLFN13)
ggplot(clink.and.rna, aes(SLFN13)) + geom_histogram(bins=20)
```

```{r}
clink.and.rna.amounts$SLFN13_amount <-
  cut(clink.and.rna.amounts$SLFN13,
      breaks=c(-Inf, 3.8, Inf),
      labels=c("low", "high"))

ggsurvplot(survfit(Surv(times, patient.vital_status)~SLFN13_amount,
                   data=clink.and.rna.amounts), pval=TRUE, conf.int=TRUE)
```

```{r}
mean(clink.and.rna$FGG)
hist(clink.and.rna$FGG)
ggplot(clink.and.rna, aes(FGG)) + geom_histogram(bins=20)
```

```{r}
clink.and.rna.amounts$FGG_amount <-
  cut(clink.and.rna.amounts$FGG,
      breaks=c(-Inf, 1.7, Inf),
      labels=c("low", "high"))

ggsurvplot(survfit(Surv(times, patient.vital_status)~FGG_amount,
                   data=clink.and.rna.amounts), pval=TRUE, conf.int=TRUE)
```

```{r}
hazards <- survival.coxph %>%
  filter(HR > 1 & Pval < 0.05) %>%
  arrange(desc(HR))
writeLines(rownames(hazards), "surv_hazards.txt")
dim(hazards)

protectors <- survival.coxph %>%
  filter(HR < 1 & Pval < 0.05) %>%
  arrange(HR)
writeLines(rownames(protectors), "surv_protectors.txt")
dim(protectors)
```

если будет ранжирование, то можно будет делать пасвэй анализ

```{r}
library(fgsea)
load("keggSymbolHuman.rdata")
```

FGSEA

```{r}
stats <- survival.coxph$HR
names(stats) <- rownames(survival.coxph)

fgseaResults_kegg <- fgseaMultilevel(keggSymbolHuman, stats, minSize = 15, maxSize = 500)
head(fgseaResults_kegg %>% arrange(padj))
```

```{r}
topPathwaysUp <- fgseaResults_kegg[ES > 0, ][head(order(padj), n=5), pathway]
topPathwaysDown <- fgseaResults_kegg[ES < 0, ][head(order(padj), n=5), pathway]
topPathways <- c(topPathwaysUp, rev(topPathwaysDown))
plotGseaTable(keggSymbolHuman[topPathways], stats, fgseaResults_kegg, gseaParam = 0.5,
              pathwayLabelStyle=list(size=8))
```

msigdb hallmarks pathways enrichment

```{r}
library(msigdb)
```

```{r}
msigdb.hs = getMsigdb(org = 'hs',id = c("SYM", "EZID"))
# Downloading and integrating KEGG gene sets
msigdb.hs = appendKEGG(msigdb.hs)
length(msigdb.hs)
```

```{r}
# You can select a child gene set according to the requirements
listCollections(msigdb.hs)
subset_msigdb = subsetCollection(msigdb.hs, c('h', 'c2', 'c5'))

#msigdb_ids <- geneIds(msigdb.hs)
msigdb_ids <- geneIds(subset_msigdb)
```

```{r}
fgseaResults_msigdb <- fgseaMultilevel(msigdb_ids, stats, minSize = 15, maxSize = 500)
head(fgseaResults_msigdb %>% arrange(padj))
```

I guess I have found good results

```{r}
topPathwaysUp <- fgseaResults_msigdb[ES > 0, ][head(order(padj), n=5), pathway]
topPathwaysDown <- fgseaResults_msigdb[ES < 0, ][head(order(padj), n=5), pathway]
topPathways <- c(topPathwaysUp, rev(topPathwaysDown))
plotGseaTable(msigdb_ids[topPathways], stats, fgseaResults_msigdb, gseaParam = 0.5,
              pathwayLabelStyle=list(size=5))
```

```{r}
clinik.and.rna.pathways <- clink.and.rna
```

```{r}
for(pathway in topPathways) {
  genes <- intersect(unique(unlist(msigdb_ids[pathway])), colnames(clinik.and.rna.pathways))

  clinik.and.rna.pathways[pathway] <- rowMeans(clinik.and.rna.pathways[, genes])
}

pathways_to_test <- topPathways

for(i in 1:10) {
  genes <- sample(rownames(counts.rna.voom), 300)
  clinik.and.rna.pathways[paste("pathway_", i, sep="")] <- rowMeans(clinik.and.rna.pathways[, genes])
  pathways_to_test <- c(pathways_to_test, paste("pathway_", i, sep=""))
}

head(clinik.and.rna.pathways[pathways_to_test])
#clinik.and.rna[, test_pathway[, 1][test_pathway[, 1] %in% colnames(clinik.and.rna)]]
#rowMeans(clinik.and.rna[, test_pathway[, 1][test_pathway[, 1] %in% colnames(clinik.and.rna)]])
```

```{r}
Survfun <- function(gene) as.formula(paste("Surv(times,patient.vital_status)",gene,sep="~"))

survival.coxph.pathways <- 
  sapply(pathways_to_test, function(gene) summary(coxph(Survfun(gene), data=clinik.and.rna.pathways)))
```

```{r}
survival.coxph.pathways <- survival.coxph.pathways %>% t() %>% as.data.frame() %>% dplyr::select(coefficients)
survival.coxph.pathways[1,]
coxph(Surv(times,patient.vital_status)~LINDGREN_BLADDER_CANCER_CLUSTER_2B, data=clinik.and.rna.pathways)
```

```{r}
survival.coxph.pathways.rownames <- rownames(survival.coxph.pathways)

survival.coxph.pathways <- as.data.frame(matrix(unlist(survival.coxph.pathways),ncol=5,byrow=TRUE))
rownames(survival.coxph.pathways) <- survival.coxph.pathways.rownames
colnames(survival.coxph.pathways) <- c("coef", "HR", "se", "z", "Pval")

survival.coxph.pathways %>% arrange(Pval)
```

```{r}
mean(clinik.and.rna.pathways$BLANCO_MELO_BRONCHIAL_EPITHELIAL_CELLS_INFLUENZA_A_DEL_NS1_INFECTION_UP)
hist(clinik.and.rna.pathways$BLANCO_MELO_BRONCHIAL_EPITHELIAL_CELLS_INFLUENZA_A_DEL_NS1_INFECTION_UP)
ggplot(clinik.and.rna.pathways, aes(BLANCO_MELO_BRONCHIAL_EPITHELIAL_CELLS_INFLUENZA_A_DEL_NS1_INFECTION_UP)) + geom_histogram(bins=20)
```

```{r}
clinik.and.rna.pathways$BLANCO_MELO_BRONCHIAL_EPITHELIAL_CELLS_INFLUENZA_A_DEL_NS1_INFECTION_UP_amount <-
  cut(clinik.and.rna.pathways$BLANCO_MELO_BRONCHIAL_EPITHELIAL_CELLS_INFLUENZA_A_DEL_NS1_INFECTION_UP,
      breaks=c(-Inf, 5.177362, Inf),
      labels=c("low", "high"))

ggsurvplot(survfit(Surv(times, patient.vital_status)~BLANCO_MELO_BRONCHIAL_EPITHELIAL_CELLS_INFLUENZA_A_DEL_NS1_INFECTION_UP_amount,
                   data=clinik.and.rna.pathways), pval=TRUE, conf.int=TRUE)
```

```{r}
mean(clinik.and.rna.pathways$RODWELL_AGING_KIDNEY_UP)
hist(clinik.and.rna.pathways$RODWELL_AGING_KIDNEY_UP)
ggplot(clinik.and.rna.pathways, aes(RODWELL_AGING_KIDNEY_UP)) + geom_histogram(bins=20)
```

```{r}
clinik.and.rna.pathways$RODWELL_AGING_KIDNEY_UP_amount <-
  cut(clinik.and.rna.pathways$RODWELL_AGING_KIDNEY_UP,
      breaks=c(-Inf, 5.979445, Inf),
      labels=c("low", "high"))

ggsurvplot(survfit(Surv(times, patient.vital_status)~RODWELL_AGING_KIDNEY_UP_amount,
                   data=clinik.and.rna.pathways), pval=TRUE, conf.int=TRUE)
```

```{r}
mean(clinik.and.rna.pathways$SMID_BREAST_CANCER_NORMAL_LIKE_UP)
hist(clinik.and.rna.pathways$SMID_BREAST_CANCER_NORMAL_LIKE_UP)
ggplot(clinik.and.rna.pathways, aes(SMID_BREAST_CANCER_NORMAL_LIKE_UP)) + geom_histogram(bins=20)
```

```{r}
clinik.and.rna.pathways$SMID_BREAST_CANCER_NORMAL_LIKE_UP_amount <-
  cut(clinik.and.rna.pathways$SMID_BREAST_CANCER_NORMAL_LIKE_UP,
      breaks=c(-Inf, 4.874391, Inf),
      labels=c("low", "high"))

ggsurvplot(survfit(Surv(times, patient.vital_status)~SMID_BREAST_CANCER_NORMAL_LIKE_UP_amount,
                   data=clinik.and.rna.pathways), pval=TRUE, conf.int=TRUE)
```

```{r}
mean(clinik.and.rna.pathways$LINDGREN_BLADDER_CANCER_CLUSTER_2B)
hist(clinik.and.rna.pathways$LINDGREN_BLADDER_CANCER_CLUSTER_2B)
ggplot(clinik.and.rna.pathways, aes(LINDGREN_BLADDER_CANCER_CLUSTER_2B)) + geom_histogram(bins=20)
```

```{r}
clinik.and.rna.pathways$LINDGREN_BLADDER_CANCER_CLUSTER_2B_amount <-
  cut(clinik.and.rna.pathways$LINDGREN_BLADDER_CANCER_CLUSTER_2B,
      breaks=c(-Inf, 5.558428, Inf),
      labels=c("low", "high"))

ggsurvplot(survfit(Surv(times, patient.vital_status)~LINDGREN_BLADDER_CANCER_CLUSTER_2B_amount,
                   data=clinik.and.rna.pathways), pval=TRUE, conf.int=TRUE)
```
