---
title: "Pathways from msigdb to Coxp"
output: html_notebook
---

```{r warning=FALSE, message=FALSE}
library(dplyr) # loads dplyr
library(survival) # core survival analysis functions
library(survminer) # recommended for visualizing survival curves
# Bioconductor packages
library(RTCGA)
library(RTCGA.clinical)
library(RTCGA.mRNA)
set.seed(1)
```

# Собираем данные

clinical data

```{r}
clinkLung <- survivalTCGA(LUSC.clinical,
                          extract.cols=c("admin.disease_code",
                                         "patient.gender",
                                         "patient.days_to_death",
                                         "patient.days_to_last_followup"))

head(clinkLung)

hist(clinkLung$times)
```

RNA

```{r}
raw.counts <- readRDS("LUSC_counts.rds")
dim(raw.counts)
as.data.frame(raw.counts[1:5, 1:5])
```

```{r}
dim(raw.counts)
counts.rna <- raw.counts[,!duplicated(substr(colnames(raw.counts), 1, 12))]
dim(counts.rna)
colnames(counts.rna) <- substr(colnames(counts.rna), 1, 12)
patients.rows <- Reduce(intersect,list(clinkLung$bcr_patient_barcode,colnames(counts.rna)))
counts.rna <- counts.rna %>% 
  as.data.frame() %>%
  dplyr::select(order(colnames(counts.rna))) %>%
  dplyr::select(patients.rows) %>%
  dplyr::slice_max(rowSums(.), n = 10000)
dim(counts.rna)
counts.rna[1:5, 1:5]
```

для cox regression надо, чтобы данные были в лог-шкале и нормализованные между образцами
нормализацию можно через limma::voom, можно через rlog (но это вероятно будет долго), можно edgeR TMM и log после этого

```{r}
counts.rna.voom <- limma::voom(counts.rna, plot = T)

#head(as.data.frame(counts.rna.voom$targets))
#head(as.data.frame(counts.rna.voom$weights))
#head(as.data.frame(counts.rna.voom$design))
#head(as.data.frame(counts.rna.voom$E))
boxplot(as.data.frame(counts.rna.voom[, 1:10]$E))

#counts.rna.log.norm <- limma::normalizeBetweenArrays(log2(counts.rna + 1))
#boxplot(counts.rna.log.norm[, 1:10])
```

concatenate rna + clinical data

```{r}
clink.and.rna <- as.data.frame(t(counts.rna.voom$E))
clink.and.rna$bcr_patient_barcode <- substr(rownames(clink.and.rna), 1, 12)
clink.and.rna <- clink.and.rna %>%
  inner_join(clinkLung, by="bcr_patient_barcode")
dim(clink.and.rna)
head(clink.and.rna)
```

msigdb hallmarks pathways enrichment

```{r}
library(msigdb)
```

```{r}
msigdb.hs = getMsigdb(org = 'hs',id = c("SYM", "EZID"))
# Downloading and integrating KEGG gene sets
msigdb.hs = appendKEGG(msigdb.hs)
length(msigdb.hs)
```

```{r}
# You can select a child gene set according to the requirements
listCollections(msigdb.hs)
subset_msigdb = subsetCollection(msigdb.hs, c('h', 'c2', 'c5'))

#msigdb_ids <- geneIds(msigdb.hs)
msigdb_ids <- geneIds(subset_msigdb)
```

```{r}
library(stringr)
```

```{r}
clinik.and.rna.pathways <- clink.and.rna

pathways_to_test <-
  unlist(lapply(names(msigdb_ids),
                function(pathway) str_replace_all(pathway, "[^[[:alnum:]|_]]", "")))
# writeLines(pathways_to_test, "pathways.txt")
```

started at:
"2023-06-22 08:51:57 MSK"
finished at:
"2023-06-22 09:16:41 MSK"

code below works about 30 minutes)

```{r}
Sys.time()

Survfun <- function(gene) as.formula(paste("Surv(times,patient.vital_status)",gene,sep="~"))

survival.coxph.pathways <- data.frame(
  coef=double(),
  HR=double(),
  se=double(),
  z=double(),
  Pval=double(), #"Pr(>|z|)"
  stringsAsFactors=F)

for(i in c(1:length(pathways_to_test))) {
  pathway <- pathways_to_test[i]
  print(i)
  genes <- intersect(unique(unlist(msigdb_ids[i])), colnames(clinik.and.rna.pathways))
  clinik.and.rna.pathways.tmp <- clink.and.rna
  
  if (length(genes) > 0) {
    if (length(genes) == 1) {
      clinik.and.rna.pathways.tmp[pathway] <- clinik.and.rna.pathways[, genes]
    } else {
     clinik.and.rna.pathways.tmp[pathway] <- rowMeans(clinik.and.rna.pathways[, genes])
    }
  
    survival.coxph.pathway <-
      as.data.frame(summary(coxph(Survfun(pathway), data=clinik.and.rna.pathways.tmp))$coefficients)
    colnames(survival.coxph.pathway) <- c("coef", "HR", "se", "z", "Pval")
  
    survival.coxph.pathways <- rbind(survival.coxph.pathways, survival.coxph.pathway)
  }
}

dim(survival.coxph.pathways)
head(survival.coxph.pathways %>% arrange(Pval), 10)

Sys.time()
```
